---
title: "Case 2 - Portfolio Management and Risk Consultation"
output: html_document
date: "2026-08-14"
---
## Group Alpha

- Aragon, Enrique Miguel Moreno
- Cardiño, Joaquin Carlo Esteban
- Carlos, Jay Michael Tenorio
- Chua, Erin Gabrielle Tan

This is the code associated with a consulting report on a portfolio 
containing a mixture of assets, specifically the Philippine stocks BDO, ICT, and SM,
the international stocks AAPL and MSFT, and the cryptocurrency BTC. We use SPY
as the primary benchmark to compare against.

# Setup

Initialize packages.

```{r}
pacman::p_load(dplyr, ggplot2, quantmod, tidyverse, tidyquant, lubridate, readxl, timetk, zoo, httr, jsonlite, quadprog)
```

# Data Acquisition

First, we pull the ticker data from Yahoo! Finance and PSE Edge.

```{r}
DATE_FROM <- "2020-01-02"
DATE_FROM_MDY <- "01-02-2020"

DATE_TO <- "2025-12-31"
DATE_TO_MDY <- "12-31-2025"
```

Many PSE stocks are no longer offered on Yahoo! Finance, so we use another free
alternative called PSE Edge.

```{r}
pse_pull <- function(cmpy_id, security_id, start = DATE_FROM_MDY, end = DATE_TO_MDY) {
  resp <- POST("https://edge.pse.com.ph/common/DisclosureCht.ax",
               body = list(cmpy_id = cmpy_id, security_id = security_id,
                           startDate = start, endDate = end),   # MM-DD-YYYY
               encode = "json",
               add_headers(
                 `X-Requested-With` = "XMLHttpRequest",
                 Referer = "https://edge.pse.com.ph/companyPage/stockData.do",
                 Origin = "https://edge.pse.com.ph",
                 Accept = "application/json, text/javascript, */*; q=0.01"))
  df <- fromJSON(content(resp, "text", encoding = "UTF-8"))$chartData
  df$date <- as.Date(df$CHART_DATE, format = "%b %d, %Y 00:00:00")
  df <- distinct(df, date, .keep_all = TRUE)
  df[order(df$date), c("date", "OPEN", "HIGH", "LOW", "CLOSE")]
}

yahoo_pull <- function(sym, start = DATE_FROM, end = DATE_TO) {
  tq_get(sym, from = start, to = end) |>
    transmute(date = as.Date(date), !!sym := round(adjusted, 4))
}

Sys.sleep(1); bdo <- pse_pull(260, 468); Sys.sleep(1)
Sys.sleep(1); ict <- pse_pull(83, 142);  Sys.sleep(1)
Sys.sleep(1); sm  <- pse_pull(599, 520); Sys.sleep(1)

# Pull USD/PHP exchange rate
fx <- yahoo_pull("USDPHP=X", start = DATE_FROM, end = DATE_TO) |> 
  rename(USDPHP = `USDPHP=X`)

prices <- list(
  BDO = bdo |> rename(BDO = CLOSE),
  ICT = ict |> rename(ICT = CLOSE),
  SM  = sm  |> rename(SM = CLOSE),
  AAPL = yahoo_pull("AAPL"), MSFT = yahoo_pull("MSFT"), BTC = yahoo_pull("BTC-USD"), SPY = yahoo_pull("SPY")
) |> reduce(inner_join, by = "date") |> arrange(date)

# Combine and convert USD assets to PHP
prices_php <- list(
  BDO = bdo |> rename(BDO = CLOSE),
  ICT = ict |> rename(ICT = CLOSE),
  SM  = sm  |> rename(SM = CLOSE),
  AAPL = yahoo_pull("AAPL"),
  MSFT = yahoo_pull("MSFT"),
  BTC  = yahoo_pull("BTC-USD"),
  SPY  = yahoo_pull("SPY")
) |> 
  reduce(inner_join, by = "date") |> 
  inner_join(fx, by = "date") |> 
  mutate(
    AAPL_PHP = AAPL * USDPHP,
    MSFT_PHP = MSFT * USDPHP,
    BTC_PHP  = `BTC-USD`  * USDPHP,
    SPY_PHP  = SPY  * USDPHP
  ) |> 
  arrange(date)

dir.create("data", showWarnings = FALSE)
write.csv(prices_php, "data/prices.csv", row.names = FALSE)
save(prices_php, file = "data/prices.RData")
```

We also get the 91-day Treasury bill as the risk-free rate.
We have to manually download it from https://data.bettergov.ph/datasets/12/resources/159
because the website blocks automated downloads.

```{r}
tbill91 <- unlist(lapply(2020:2025, function(yr) {
  gs <- read_excel("data/gs_rates_monthly.xlsx", sheet = as.character(yr), skip = 4)
  row_91 <- gs[which(grepl("91-day", as.character(gs[[2]]))), ]   # label in col 2
  as.numeric(unlist(row_91[1, seq(3, 25, by = 2)]))               # C,E,G,...,Y
}))
```

We then clean the Treasury bill rates by dropping the blank months and
computing the annual and daily risk-free rate.

```{r}
tbill91_pct <- na.omit(tbill91)                # drops blank / no-auction months
rf_annual <- mean(tbill91_pct) / 100           # ~4% expected
rf_daily  <- (1 + rf_annual)^(1 / 252) - 1     # use in Sharpe / frontier
```

# Exploratory Financial Analysis

First we compute for daily log returns to get a simple overview of each asset's
performance.

```{r}
prices_php <- mutate(prices_php,
                 BDO_return = log(BDO / lag(BDO, 1)),
                 ICT_return = log(ICT / lag(ICT, 1)),
                 SM_return = log(SM / lag(SM, 1)),
                 AAPL_return = log(AAPL / lag(AAPL, 1)),
                 MSFT_return = log(MSFT / lag(MSFT, 1)),
                 BTC_return = log(`BTC-USD` / lag(`BTC-USD`, 1)),
                 SPY_return = log(SPY / lag(SPY, 1)))
```

Then present the findings in a table for easier understanding.

```{r}
findings_summary <- prices_php |>
  select(date, BDO, ICT, SM, AAPL, MSFT, `BTC-USD`, SPY) |>
  pivot_longer(-date, names_to = "asset", values_to = "price") |>
  group_by(asset) |>
  mutate(ret = log(price / lag(price))) |>
  summarise(
    `Mean Daily Ret (%)` = mean(ret, na.rm = TRUE) * 100,
    `SD Daily Ret (%)`   = sd(ret, na.rm = TRUE) * 100,
    `Ann. Ret (%)`       = mean(ret, na.rm = TRUE) * 252 * 100,
    `Ann. Vol (%)`       = sd(ret, na.rm = TRUE) * sqrt(252) * 100,
    Sharpe               = (mean(ret, na.rm = TRUE) * 252 - rf_annual) /
                           (sd(ret, na.rm = TRUE) * sqrt(252)),
    `Max Drawdown (%)`   = min(cumprod(1 + ret) / cummax(cumprod(1 + ret)) - 1) * 100
  ) |>
  ungroup()


knitr::kable(findings_summary, digits = 3)
```

# Portfolio Construction

We construct a portfolio using these 6 assets: BDO, ICT, SM, AAPL, MSFT, and BTC. 
We start with building a returns matrix as each optimization technique (Efficient Frontier, 
Global Minimum Variance, and Maximum Sharpe) relies on return data.

```{r}
ret_data <- prices_php |>
  select(date, ends_with("_return"), -SPY_return) |>
  na.omit()

assets <- c("BDO", "ICT", "SM", "AAPL", "MSFT", "BTC")

returns <- as.matrix(select(ret_data, ends_with("_return")))
colnames(returns) <- assets

n_assets <- length(assets)

print(n_assets)
print(head(returns))
```

We annualize the mean return vector and covariance matrix, since all our
optimization and risk targets are expressed on an annual basis.

```{r}
#expected return vector
mu <- colMeans(returns) * 252

# risk
Sigma <- cov(returns) * 252

data.frame(Asset = assets, `Annual Return (%)` = round(mu * 100, 2), check.names = FALSE)
```

To see the effects of optimization on our portfolio, we create a baseline of a 
a simple equal weighted portfolio, with anf without rebalancing. This baseline gives us something to 
compare our optimized portfolios against.

```{r}
portfolio_growth <- function(weights, ret_df = ret_data, rebalance = FALSE) {
  dates <- ret_df$date
  rets <- as.matrix(select(ret_df, ends_with("_return")))

  if (!rebalance) {
    asset_growth <- apply(rets, 2, function(r) cumprod(1 + r))
    port_value <- as.numeric(asset_growth %*% weights)
  } else {
    month_id <- format(dates, "%Y-%m")
    port_value <- numeric(length(dates))
    value <- 1
    holdings <- value * weights

    for (i in seq_along(dates)) {
      if (i > 1 && month_id[i] != month_id[i - 1]) {
        holdings <- value * weights 
      }
      holdings <- holdings * (1 + rets[i, ])
      value <- sum(holdings)
      port_value[i] <- value
    }
  }

  tibble(date = dates, value = port_value)
}
```

We rebalance because prices move and starts pulling the actual portfolio away from that optimum.
Over time, some assets grow faster than others which can shift the target mix and exposes the assets to 
more risk than planned.

```{r}
eq_weights <- rep(1 / n_assets, n_assets)

eq_no_rebal <- portfolio_growth(eq_weights, rebalance = FALSE) |> mutate(portfolio = "Equal Weight (No Rebalance)")
eq_rebal <- portfolio_growth(eq_weights, rebalance = TRUE)  |> mutate(portfolio = "Equal Weight (Monthly Rebalance)")

```

Lastly, we initialize equal weighing among the assets and call `portfolio_growth()` to create paths with and without monthly rebalancing.

# Portfolio Optimization

With the return matrix and baseline established, we now solve for the three portfolios, the Efficient Frontier, the Global
Minimum Variance (GMV) Portfolio, and the Maximum Sharpe Portfolio. We use the `quadprog` (quadratic programming) library to help with portfolio optimization. In general, these 3 techniques or theories all aim to achieve a singular goal, that being how to best balance or optimze the ratio
of returns given risks in its own way.

### Efficient Frontier

The efficient frontier is a core concept in Modern Portfolio Theory (MPT) introduced by Harry Markowitz in 1952. It is a curved line on a graph that shows the best investment portfolios, these portfolios give the highest expected return for a set level of risk, or the lowest risk for a set level of return.

We trace the frontier by minimizing variance for a grid of target returns spanning the achievable range.

```{r}
target_rets <- seq(min(mu), max(mu), length.out = 100)

eff_frontier <- map_dfr(target_rets, function(tr) {
  Amat_ef <- cbind(rep(1, n_assets), mu, diag(n_assets))
  bvec_ef <- c(1, tr, rep(0, n_assets))

  sol <- tryCatch(solve.QP(Dmat, dvec, Amat_ef, bvec_ef, meq = 2),
                   error = function(e) NULL)
  if (is.null(sol)) return(NULL)

  w <- sol$solution
  tibble(
    target_ret = tr,
    ret = sum(w * mu),
    vol = as.numeric(sqrt(t(w) %*% Sigma %*% w)),
    sharpe = (sum(w * mu) - rf_annual) / as.numeric(sqrt(t(w) %*% Sigma %*% w)),
    weights = list(setNames(w, assets))
  )
})

target_rets
```

### Global Minimum Variance Portfolio

A Global Minimum Variance (GMV) Portfolio is a specialized asset mix built from a set of risky investments that achieves the lowest possible total risk or price volatility (variance). It sits at the absolute leftmost point of the efficient frontier curve in MPT.

The GMV portfolio minimizes `w'Σw` subject to weights summing to 1. 
```{r}
Dmat <- 2 * Sigma
dvec <- rep(0, n_assets)

Amat <- cbind(rep(1, n_assets), diag(n_assets))   # sum(w) = 1, w_i >= 0
bvec <- c(1, rep(0, n_assets))

gmv_sol <- solve.QP(Dmat, dvec, Amat, bvec, meq = 1)
gmv_weights <- setNames(gmv_sol$solution, assets)

gmv_ret <- sum(gmv_weights * mu)
gmv_vol <- as.numeric(sqrt(t(gmv_weights) %*% Sigma %*% gmv_weights))

knitr::kable(data.frame(Asset = assets, Weight = round(gmv_weights, 4)))
```

### Maximum Sharpe Portfolio
A Maximum Sharpe Portfolio is the optimal asset allocation on the efficient frontier that delivers the highest possible excess return per unit of total risk (volatility).

```{r}
max_sharpe_row <- eff_frontier |> slice_max(sharpe, n = 1)
max_sharpe_weights <- max_sharpe_row$weights[[1]]

knitr::kable(data.frame(Asset = assets, Weight = round(max_sharpe_weights, 4)))
```


# Portfolio Risk Assessment


# Executive Recommendation


# References
