---
title: "Case 2 - Portfolio Management and Risk Consultation"
output: html_document
date: "2026-08-14"
---
## Group Alpha

- Aragon, Enrique Miguel Moreno
- Cardiño, Joaquin Carlo Esteban
- Carlos, Jay Michael Tenorio
- Chua, Erin Gabrielle Tan

This is the code associated with a consulting report on a portfolio 
containing a mixture of assets, specifically the Philippine stocks BDO, ICT, and SM,
the international stocks AAPL and MSFT, and the cryptocurrency BTC. We use SPY
as the primary benchmark to compare against.

# Setup

Initialize packages.

```{r}
pacman::p_load(dplyr, ggplot2, quantmod, tidyverse, tidyquant, lubridate, readxl, timetk, zoo, httr, jsonlite, quadprog, PerformanceAnalytics, xts, flextable)
```

```{r}
RANDOM_SEED <- 42
```

# Data Acquisition

First, we pull the ticker data from Yahoo! Finance and PSE Edge.
```{r}
DATE_FROM <- "2020-01-02"
DATE_FROM_MDY <- "01-02-2020"

DATE_TO <- "2025-12-31"
DATE_TO_MDY <- "12-31-2025"
```

Many PSE stocks are no longer offered on Yahoo! Finance, so we use another free
alternative called PSE Edge.

```{r}
pse_pull <- function(cmpy_id, security_id, start = DATE_FROM_MDY, end = DATE_TO_MDY) {
  resp <- POST("https://edge.pse.com.ph/common/DisclosureCht.ax",
               body = list(cmpy_id = cmpy_id, security_id = security_id,
                           startDate = start, endDate = end),   # MM-DD-YYYY
               encode = "json",
               add_headers(
                 `X-Requested-With` = "XMLHttpRequest",
                 Referer = "https://edge.pse.com.ph/companyPage/stockData.do",
                 Origin = "https://edge.pse.com.ph",
                 Accept = "application/json, text/javascript, */*; q=0.01"))
  df <- fromJSON(content(resp, "text", encoding = "UTF-8"))$chartData
  df$date <- as.Date(df$CHART_DATE, format = "%b %d, %Y 00:00:00")
  df <- distinct(df, date, .keep_all = TRUE)
  df[order(df$date), c("date", "CLOSE")]
}

yahoo_pull <- function(sym, start = DATE_FROM, end = DATE_TO) {
  tq_get(sym, from = start, to = end) |>
    transmute(date = as.Date(date), !!sym := round(adjusted, 4))
}

Sys.sleep(1); bdo <- pse_pull(260, 468); Sys.sleep(1)
Sys.sleep(1); ict <- pse_pull(83, 142);  Sys.sleep(1)
Sys.sleep(1); sm  <- pse_pull(599, 520); Sys.sleep(1)

# Pull USD/PHP exchange rate
fx <- yahoo_pull("USDPHP=X", start = DATE_FROM, end = DATE_TO) |> 
  rename(USDPHP = `USDPHP=X`)

# Combine and convert USD assets to PHP
prices_php <- list(
  BDO = bdo |> rename(BDO = CLOSE),
  ICT = ict |> rename(ICT = CLOSE),
  SM  = sm  |> rename(SM = CLOSE),
  AAPL = yahoo_pull("AAPL"),
  MSFT = yahoo_pull("MSFT"),
  BTC  = yahoo_pull("BTC-USD"),
  SPY  = yahoo_pull("SPY")
) |> 
  reduce(inner_join, by = "date") |> 
  inner_join(fx, by = "date") |> 
  mutate(
    AAPL_PHP = AAPL * USDPHP,
    MSFT_PHP = MSFT * USDPHP,
    BTC_PHP  = `BTC-USD`  * USDPHP,
    SPY_PHP  = SPY  * USDPHP
  ) |> 
  arrange(date)

# Save a clean, peso-denominated data file for the deliverable
prices_clean <- prices_php |>
  select(date, USDPHP, BDO, ICT, SM, AAPL_PHP, MSFT_PHP, BTC_PHP, SPY_PHP)

dir.create("data", showWarnings = FALSE)
write.csv(prices_clean, "data/prices.csv", row.names = FALSE)
save(prices_clean, file = "data/prices.RData")
```

We also get the 91-day Treasury bill as the risk-free rate.
We have to manually download it from https://data.bettergov.ph/datasets/12/resources/159
because the website blocks automated downloads.

```{r}
tbill91 <- unlist(lapply(2020:2025, function(yr) {
  gs <- read_excel("data/gs_rates_monthly.xlsx", sheet = as.character(yr), skip = 4)
  row_91 <- gs[which(grepl("91-day", as.character(gs[[2]]))), ]   # label in col 2
  as.numeric(unlist(row_91[1, seq(3, 25, by = 2)]))               # C,E,G,...,Y
}))
```

We then clean the Treasury bill rates by dropping the blank months and
computing the annual and daily risk-free rate.

```{r}
tbill91_pct <- na.omit(tbill91)                # drops blank / no-auction months
rf_annual <- mean(tbill91_pct) / 100           # ~4% expected
rf_daily  <- (1 + rf_annual)^(1 / 252) - 1     # use in Sharpe / frontier
```

# Exploratory Financial Analysis

First we compute for daily log returns to get a simple overview of each asset's
performance.

```{r}
prices_php <- prices_php |>
              mutate(
                BDO_return  = log(BDO / lag(BDO)),
                ICT_return  = log(ICT / lag(ICT)),
                SM_return   = log(SM / lag(SM)),
                AAPL_return = log(AAPL_PHP / lag(AAPL_PHP)),
                MSFT_return = log(MSFT_PHP / lag(MSFT_PHP)),
                BTC_return  = log(BTC_PHP / lag(BTC_PHP)),
                SPY_return  = log(SPY_PHP / lag(SPY_PHP))
              )
```

Then present the findings in a table for easier understanding.

```{r}
rets_wide <- prices_php |> select(date, ends_with("_return")) |> na.omit()

findings_summary <- rets_wide |>
  pivot_longer(-date, names_to = "asset", values_to = "ret") |>
  group_by(asset) |>
  summarise(
    `Mean Daily Ret (%)` = mean(ret, na.rm = TRUE) * 100,
    `SD Daily Ret (%)`   = sd(ret, na.rm = TRUE) * 100,
    `Ann. Ret (%)`       = mean(ret, na.rm = TRUE) * 252 * 100,
    `Ann. Vol (%)`       = sd(ret, na.rm = TRUE) * sqrt(252) * 100,
    Sharpe               = (mean(ret, na.rm = TRUE) * 252 - rf_annual) /
                           (sd(ret, na.rm = TRUE) * sqrt(252)),
    `Max Drawdown (%)` = { r <- ret[!is.na(ret)];
                       min(cumprod(1 + r) / cummax(cumprod(1 + r)) - 1) * 100 }
  ) |>
  ungroup()


print(findings_summary)
```


## Descriptive Statistics

We create a table comprised of the returns of each asset to compute for descriptive statistics.

```{r}
returns_tbl <- prices_php  |> 

  select(date,
         BDO_return,
         ICT_return,
         SM_return,
         AAPL_return,
         MSFT_return,
         BTC_return,
         SPY_return)

head(returns_tbl)
dim(returns_tbl)
```

```{r}
descriptive_table <- tibble(

  Statistic = c(
    "Mean",
    "Median",
    "Trimmed Mean",
    "Standard Deviation",
    "Variance",
    "Minimum",
    "Q1",
    "Q3",
    "Maximum",
    "Range",
    "Interquartile Range",
    "Coefficient of Variation",
    "Skewness",
    "Kurtosis",
    "Observations"
  ),

  BDO = c(
    mean(returns_tbl$BDO_return, na.rm = TRUE),
    median(returns_tbl$BDO_return, na.rm = TRUE),
    mean(returns_tbl$BDO_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$BDO_return, na.rm = TRUE),
    var(returns_tbl$BDO_return, na.rm = TRUE),
    min(returns_tbl$BDO_return, na.rm = TRUE),
    quantile(returns_tbl$BDO_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$BDO_return, 0.75, na.rm = TRUE),
    max(returns_tbl$BDO_return, na.rm = TRUE),
    diff(range(returns_tbl$BDO_return, na.rm = TRUE)),
    IQR(returns_tbl$BDO_return, na.rm = TRUE),
    sd(returns_tbl$BDO_return, na.rm = TRUE) /
      mean(returns_tbl$BDO_return, na.rm = TRUE),
    skewness(returns_tbl$BDO_return, na.rm = TRUE),
    kurtosis(returns_tbl$BDO_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$BDO_return))
  ),
  
  ICT = c(
    mean(returns_tbl$ICT_return, na.rm = TRUE),
    median(returns_tbl$ICT_return, na.rm = TRUE),
    mean(returns_tbl$ICT_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$ICT_return, na.rm = TRUE),
    var(returns_tbl$ICT_return, na.rm = TRUE),
    min(returns_tbl$ICT_return, na.rm = TRUE),
    quantile(returns_tbl$ICT_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$ICT_return, 0.75, na.rm = TRUE),
    max(returns_tbl$ICT_return, na.rm = TRUE),
    diff(range(returns_tbl$ICT_return, na.rm = TRUE)),
    IQR(returns_tbl$ICT_return, na.rm = TRUE),
    sd(returns_tbl$ICT_return, na.rm = TRUE) /
      mean(returns_tbl$ICT_return, na.rm = TRUE),
    skewness(returns_tbl$ICT_return, na.rm = TRUE),
    kurtosis(returns_tbl$ICT_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$ICT_return))
  ),
  
  SM = c(
    mean(returns_tbl$SM_return, na.rm = TRUE),
    median(returns_tbl$SM_return, na.rm = TRUE),
    mean(returns_tbl$SM_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$SM_return, na.rm = TRUE),
    var(returns_tbl$SM_return, na.rm = TRUE),
    min(returns_tbl$SM_return, na.rm = TRUE),
    quantile(returns_tbl$SM_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$SM_return, 0.75, na.rm = TRUE),
    max(returns_tbl$SM_return, na.rm = TRUE),
    diff(range(returns_tbl$SM_return, na.rm = TRUE)),
    IQR(returns_tbl$SM_return, na.rm = TRUE),
    sd(returns_tbl$SM_return, na.rm = TRUE) /
      mean(returns_tbl$SM_return, na.rm = TRUE),
    skewness(returns_tbl$SM_return, na.rm = TRUE),
    kurtosis(returns_tbl$SM_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$SM_return))
  ),
  
  AAPL = c(
    mean(returns_tbl$AAPL_return, na.rm = TRUE),
    median(returns_tbl$AAPL_return, na.rm = TRUE),
    mean(returns_tbl$AAPL_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$AAPL_return, na.rm = TRUE),
    var(returns_tbl$AAPL_return, na.rm = TRUE),
    min(returns_tbl$AAPL_return, na.rm = TRUE),
    quantile(returns_tbl$AAPL_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$AAPL_return, 0.75, na.rm = TRUE),
    max(returns_tbl$AAPL_return, na.rm = TRUE),
    diff(range(returns_tbl$AAPL_return, na.rm = TRUE)),
    IQR(returns_tbl$AAPL_return, na.rm = TRUE),
    sd(returns_tbl$AAPL_return, na.rm = TRUE) /
      mean(returns_tbl$AAPL_return, na.rm = TRUE),
    skewness(returns_tbl$AAPL_return, na.rm = TRUE),
    kurtosis(returns_tbl$AAPL_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$AAPL_return))
  ),
  
  MSFT = c(
    mean(returns_tbl$MSFT_return, na.rm = TRUE),
    median(returns_tbl$MSFT_return, na.rm = TRUE),
    mean(returns_tbl$MSFT_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$MSFT_return, na.rm = TRUE),
    var(returns_tbl$MSFT_return, na.rm = TRUE),
    min(returns_tbl$MSFT_return, na.rm = TRUE),
    quantile(returns_tbl$MSFT_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$MSFT_return, 0.75, na.rm = TRUE),
    max(returns_tbl$MSFT_return, na.rm = TRUE),
    diff(range(returns_tbl$MSFT_return, na.rm = TRUE)),
    IQR(returns_tbl$MSFT_return, na.rm = TRUE),
    sd(returns_tbl$MSFT_return, na.rm = TRUE) /
      mean(returns_tbl$MSFT_return, na.rm = TRUE),
    skewness(returns_tbl$MSFT_return, na.rm = TRUE),
    kurtosis(returns_tbl$MSFT_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$MSFT_return))
  ),
  
  BTC = c(
    mean(returns_tbl$BTC_return, na.rm = TRUE),
    median(returns_tbl$BTC_return, na.rm = TRUE),
    mean(returns_tbl$BTC_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$BTC_return, na.rm = TRUE),
    var(returns_tbl$BTC_return, na.rm = TRUE),
    min(returns_tbl$BTC_return, na.rm = TRUE),
    quantile(returns_tbl$BTC_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$BTC_return, 0.75, na.rm = TRUE),
    max(returns_tbl$BTC_return, na.rm = TRUE),
    diff(range(returns_tbl$BTC_return, na.rm = TRUE)),
    IQR(returns_tbl$BTC_return, na.rm = TRUE),
    sd(returns_tbl$BTC_return, na.rm = TRUE) /
      mean(returns_tbl$BTC_return, na.rm = TRUE),
    skewness(returns_tbl$BTC_return, na.rm = TRUE),
    kurtosis(returns_tbl$BTC_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$BTC_return))
  ),
  
  SPY = c(
    mean(returns_tbl$SPY_return, na.rm = TRUE),
    median(returns_tbl$SPY_return, na.rm = TRUE),
    mean(returns_tbl$SPY_return, trim = 0.05, na.rm = TRUE),
    sd(returns_tbl$SPY_return, na.rm = TRUE),
    var(returns_tbl$SPY_return, na.rm = TRUE),
    min(returns_tbl$SPY_return, na.rm = TRUE),
    quantile(returns_tbl$SPY_return, 0.25, na.rm = TRUE),
    quantile(returns_tbl$SPY_return, 0.75, na.rm = TRUE),
    max(returns_tbl$SPY_return, na.rm = TRUE),
    diff(range(returns_tbl$SPY_return, na.rm = TRUE)),
    IQR(returns_tbl$SPY_return, na.rm = TRUE),
    sd(returns_tbl$SPY_return, na.rm = TRUE) /
      mean(returns_tbl$SPY_return, na.rm = TRUE),
    skewness(returns_tbl$SPY_return, na.rm = TRUE),
    kurtosis(returns_tbl$SPY_return, na.rm = TRUE),
    sum(!is.na(returns_tbl$SPY_return))
  )
)

descriptive_table


```
Correlation Matrix

```{r}
correlation_matrix <- returns_tbl  |> 
  select(-date)  |> 
  cor(use = "complete.obs")
correlation_matrix

```
Covariation Matrix
```{r}
covariance_matrix <- returns_tbl |> 
  select(-date)  |> 
  cov(use = "complete.obs")
covariance_matrix

View(prices_php)
```

## Data Visualization

Figure 1: Indexed Price Trend 
```{r}
indexed_prices <- prices_php|>
  select(date, BDO, ICT, SM, AAPL_PHP, MSFT_PHP, BTC_PHP, SPY_PHP) |>
  mutate(across(-date, ~ .x / first(.x) * 100)) |>
  tidyr::pivot_longer(-date, names_to = "asset", values_to = "indexed_price")

ggplot(indexed_prices, aes(x = date, y = indexed_price, color = asset)) +
  geom_line(linewidth = 0.7, na.rm = TRUE) +
  labs(
    title = "Figure 1: Indexed Price Trend (Base = 100, Jan 2020)",
    x = "Date",
    y = "Indexed Price"
  ) +
  theme_minimal()
```
Figure 2 Daily Return Distribution
```{r}
returns_long_viz <-  prices_php|>
  select(date, BDO_return, ICT_return, SM_return, AAPL_return, MSFT_return, BTC_return, SPY_return) |>
  tidyr::pivot_longer(-date, names_to = "asset", values_to = "return")

ggplot(returns_long_viz, aes(x = return, fill = asset)) +
  geom_histogram(bins = 60) +
  facet_wrap(~asset, scales = "free") +
  labs(
    title = "Figure 2: Distribution of Daily Returns by Asset",
    x = "Daily Return",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
Figure 3 30-Day Rolling Volatility
```{r}
rolling_vol <- prices_php |>
  arrange(date) |>
  mutate(
    bdo_vol_30d = zoo::rollapply(BDO_return, width = 30, FUN = sd, fill = NA, align = "right"),
    ict_vol_30d = zoo::rollapply(ICT_return, width = 30, FUN = sd, fill = NA, align = "right"),
    sm_vol_30d = zoo::rollapply(SM_return, width = 30, FUN = sd, fill = NA, align = "right"),
    aapl_vol_30d = zoo::rollapply(AAPL_return, width = 30, FUN = sd, fill = NA, align = "right"),
    msft_vol_30d = zoo::rollapply(MSFT_return, width = 30, FUN = sd, fill = NA, align = "right"),
    btc_vol_30d = zoo::rollapply(BTC_return, width = 30, FUN = sd, fill = NA, align = "right"),
    spy_vol_30d = zoo::rollapply(SPY_return, width = 30, FUN = sd, fill = NA, align = "right")
  ) |>
  select(date, bdo_vol_30d, ict_vol_30d, sm_vol_30d, aapl_vol_30d, msft_vol_30d, btc_vol_30d, spy_vol_30d) |>
  tidyr::pivot_longer(-date, names_to = "asset", values_to = "rolling_vol")

ggplot(rolling_vol, aes(x = date, y = rolling_vol, color = asset)) +
  geom_line(linewidth = 0.6, na.rm = TRUE) +
  labs(
    title = "Figure 3: 30-Day Rolling Volatility by Asset",
    x = "Date",
    y = "30-Day Rolling Std. Dev. of Return"
  ) +
  theme_minimal()
```


## EFA

1. Comparing Results and Volatility Across Assets
```{r}
return_vol_summary <- returns_long_viz |>
  group_by(asset) |>
  summarise(
    avg_daily_return = mean(return, na.rm = TRUE),
    annualized_return = mean(return, na.rm = TRUE) * 252,
    daily_volatility = sd(return, na.rm = TRUE),
    annualized_volatility = sd(return, na.rm = TRUE) * sqrt(252)
  ) |>
  arrange(desc(annualized_return))

return_vol_summary
```

# Portfolio Construction

We construct a portfolio using these 6 assets: BDO, ICT, SM, AAPL, MSFT, and BTC. 
We start with building a returns matrix as each optimization technique (Efficient Frontier, 
Global Minimum Variance, and Maximum Sharpe) relies on return data.

```{r}
ret_data <- prices_php |>
  select(date, ends_with("_return"), -SPY_return) |>
  na.omit()

assets <- c("BDO", "ICT", "SM", "AAPL", "MSFT", "BTC")

returns <- as.matrix(select(ret_data, ends_with("_return")))
colnames(returns) <- assets

n_assets <- length(assets)

print(n_assets)
print(head(returns))
```

We annualize the mean return vector and covariance matrix, since all our
optimization and risk targets are expressed on an annual basis.

```{r}
#expected return vector
mu <- colMeans(returns) * 252

# risk
Sigma <- cov(returns) * 252

data.frame(Asset = assets, `Annual Return (%)` = round(mu * 100, 2), check.names = FALSE)
```

To see the effects of optimization on our portfolio, we create a baseline of a 
a simple equal weighted portfolio, with and without re-balancing. This baseline gives us something to 
compare our optimized portfolios against.

```{r}
portfolio_growth <- function(weights, ret_df = ret_data, rebalance = FALSE) {
  dates <- ret_df$date
  rets <- as.matrix(select(ret_df, ends_with("_return")))

  if (!rebalance) {
    asset_growth <- apply(rets, 2, function(r) cumprod(1 + r))
    port_value <- as.numeric(asset_growth %*% weights)
  } else {
    month_id <- format(dates, "%Y-%m")
    port_value <- numeric(length(dates))
    value <- 1
    holdings <- value * weights

    for (i in seq_along(dates)) {
      if (i > 1 && month_id[i] != month_id[i - 1]) {
        holdings <- value * weights 
      }
      holdings <- holdings * (1 + rets[i, ])
      value <- sum(holdings)
      port_value[i] <- value
    }
  }

  tibble(date = dates, value = port_value)
}
```

We re-balance because prices move and starts pulling the actual portfolio away from that optimum.
Over time, some assets grow faster than others which can shift the target mix and exposes the assets to 
more risk than planned.

```{r}
eq_weights <- rep(1 / n_assets, n_assets)

eq_no_rebal <- portfolio_growth(eq_weights, rebalance = FALSE) |> mutate(portfolio = "Equal Weight (No Rebalance)")
eq_rebal <- portfolio_growth(eq_weights, rebalance = TRUE)  |> mutate(portfolio = "Equal Weight (Monthly Rebalance)")

```

Lastly, we initialize equal weighing among the assets and call `portfolio_growth()` to create paths with and without monthly rebalancing.

# Portfolio Optimization

With the return matrix and baseline established, we now solve for the three portfolios, the Efficient Frontier, the Global
Minimum Variance (GMV) Portfolio, and the Maximum Sharpe Portfolio. We use the `quadprog` (quadratic programming) library to help with portfolio optimization. In general, these 3 techniques or theories all aim to achieve a singular goal, that being how to best balance or optimize the ratio
of returns given risks in its own way.

### Efficient Frontier

The efficient frontier is a core concept in Modern Portfolio Theory (MPT) introduced by Harry Markowitz in 1952. It is a curved line on a graph that shows the best investment portfolios, these portfolios give the highest expected return for a set level of risk, or the lowest risk for a set level of return.

We trace the frontier by minimizing variance for a grid of target returns spanning the achievable range.

```{r}
Dmat <- 2 * Sigma
dvec <- rep(0, n_assets)
```

```{r}
target_rets <- seq(min(mu), max(mu), length.out = 100)

eff_frontier <- map_dfr(target_rets, function(tr) {
  Amat_ef <- cbind(rep(1, n_assets), mu, diag(n_assets))
  bvec_ef <- c(1, tr, rep(0, n_assets))

  sol <- tryCatch(solve.QP(Dmat, dvec, Amat_ef, bvec_ef, meq = 2),
                   error = function(e) NULL)
  if (is.null(sol)) return(NULL)

  w <- sol$solution
  tibble(
    target_ret = tr,
    ret = sum(w * mu),
    vol = as.numeric(sqrt(t(w) %*% Sigma %*% w)),
    sharpe = (sum(w * mu) - rf_annual) / as.numeric(sqrt(t(w) %*% Sigma %*% w)),
    weights = list(setNames(w, assets))
  )
})

target_rets
```

### Global Minimum Variance Portfolio

A Global Minimum Variance (GMV) Portfolio is a specialized asset mix built from a set of risky investments that achieves the lowest possible total risk or price volatility (variance). It sits at the absolute leftmost point of the efficient frontier curve in MPT.

The GMV portfolio minimizes `w'Σw` subject to weights summing to 1. 
```{r}
Amat <- cbind(rep(1, n_assets), diag(n_assets))   # sum(w) = 1, w_i >= 0
bvec <- c(1, rep(0, n_assets))

gmv_sol <- solve.QP(Dmat, dvec, Amat, bvec, meq = 1)
gmv_weights <- setNames(gmv_sol$solution, assets)

gmv_ret <- sum(gmv_weights * mu)
gmv_vol <- as.numeric(sqrt(t(gmv_weights) %*% Sigma %*% gmv_weights))

knitr::kable(data.frame(Asset = assets, Weight = round(gmv_weights, 4)))
```

### Maximum Sharpe Portfolio
A Maximum Sharpe Portfolio is the optimal asset allocation on the efficient frontier that delivers the highest possible excess return per unit of total risk (volatility).

```{r}
max_sharpe_row <- eff_frontier |> slice_max(sharpe, n = 1)
max_sharpe_weights <- max_sharpe_row$weights[[1]]

knitr::kable(data.frame(Asset = assets, Weight = round(max_sharpe_weights, 4)))
```

We visualize the frontier together with the individual assets, the key
portfolios, and the capital allocation line through the tangency point.

```{r}
# CML slope through the tangency (max-Sharpe) portfolio
slope_cml <- (max_sharpe_row$ret - rf_annual) / max_sharpe_row$vol

# Annualized points: assets, then the key portfolios
asset_pts <- data.frame(
  asset = assets,
  ret   = colMeans(returns) * 252,
  vol   = apply(returns, 2, sd) * sqrt(252)
)

eq_ret <- as.matrix(select(ret_data, ends_with("_return"))) %*% eq_weights

key_pts <- rbind(
  asset_pts,
  data.frame(asset = "Equal Weight", ret = mean(eq_ret) * 252,  vol = sd(eq_ret) * sqrt(252)),
  data.frame(asset = "GMV",          ret = gmv_ret,             vol = gmv_vol),
  data.frame(asset = "Max Sharpe",   ret = max_sharpe_row$ret,  vol = max_sharpe_row$vol),
  data.frame(asset = "SPY (PHP)",    ret = mean(rets_wide$SPY_return) * 252,
                                     vol = sd(rets_wide$SPY_return) * sqrt(252))
)

eff_frontier |>
  arrange(vol) |>                            # safe even with NULL unfeasible gaps
  ggplot(aes(vol, ret)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_abline(intercept = rf_annual, slope = slope_cml,
              linetype = "dashed", color = "grey45") +
  geom_point(data = key_pts, aes(color = asset), size = 3.5) +
  geom_text(data = key_pts, aes(label = asset, color = asset),
            hjust = -0.1, vjust = -0.6, size = 3) +
  scale_color_viridis_d() +
  labs(title = "Efficient frontier with capital allocation line",
       subtitle = paste0("Risk-free rate = ", round(rf_annual * 100, 2), "%"),
       x = "Annualized volatility", y = "Annualized return") +
  theme_minimal() +
  coord_cartesian(xlim = c(0, max(key_pts$vol, eff_frontier$vol) * 1.25))
```

## Cross-Portfolio Visualizations

### Historical Price Plot
```{r}
# !VISUALIZATION
# 
```


### Daily Return Plot
```{r}
# !VISUALIZATION
# 
```

### Correlation Heatmap
```{r}
# !VISUALIZATION
# 
```

### Portfolio Allocation

```{r}
# !VISUALIZATION
# portfolio allocation
```

### Optimized Portfolio Growth Comparison

```{r}
# !VISUALIZATION
# portfolio growth comparison (rebalances vs no rebalance per portoflio)
```

# Portfolio Risk Assessment

Having constructed the Equal-Weight, GMV, and Maximum Sharpe portfolios, this section will assess how
each portfolio handles risk based on, 

- Risk evaluation
- Descriptive statistics per portfolio
- Visualizations

## Risk Evaluation

Here we focus on computing risk management techniques, specifically Historical VaR, Parametric VaR, Expected Shortfall (CVaR), Monte Carlo Simulation, and Stress Testing. 

We first compute a daily portfolio return series for each portfolio, this would act as the raw input for each measure. 

```{r}
eq_weights <- rep(1 / n_assets, n_assets)

port_returns <- tibble(
  date = ret_data$date,
  `Equal-Weight` = as.numeric(returns %*% eq_weights),
  `GMV` = as.numeric(returns %*% gmv_weights),
  `Max Sharpe` = as.numeric(returns %*% max_sharpe_weights)
)

#standard confidence level
conf_level <- 0.95 
```

#### Historical VaR

```{r}
historical_var <- port_returns |>
  summarise(across(
    c(`Equal-Weight`, GMV, `Max Sharpe`),
    ~ -quantile(.x, probs = 1 - conf_level, na.rm = TRUE)
  ))


data.frame(Portfolio = names(historical_var),
            `Daily Historical VaR (95%)` = round(as.numeric(historical_var), 4),
            check.names = FALSE)
```

#### Parametric VaR

```{r}
z_score <- qnorm(1 - conf_level)

parametric_var <- port_returns |>
  summarise(across(
    c(`Equal-Weight`, GMV, `Max Sharpe`),
    ~ -(mean(.x, na.rm = TRUE) + z_score * sd(.x, na.rm = TRUE))
  ))

data.frame(Portfolio = names(parametric_var),
            `Daily Parametric VaR (95%)` = round(as.numeric(parametric_var), 4),
            check.names = FALSE)

```

#### Expected Shortfall (CVaR) 

```{r}
expected_shortfall <- port_returns |>
  summarise(across(
    c(`Equal-Weight`, GMV, `Max Sharpe`),
    ~ -mean(.x[.x <= quantile(.x, probs = 1 - conf_level, na.rm = TRUE)], na.rm = TRUE)
  ))


data.frame(Portfolio = names(expected_shortfall),
            `Daily Expected Shortfall (95%)` = round(as.numeric(expected_shortfall), 4),
            check.names = FALSE)

```

#### Monte Carlo Simulation

```{r}
set.seed(RANDOM_SEED)
n_sim_val <- 5000

simulate_portfolio_var <- function(port_ret, n_sim = n_sim_val, conf = conf_level) {
  mu_p <- mean(port_ret, na.rm = TRUE)
  sigma_p <- sd(port_ret, na.rm = TRUE)

  sim_returns <- rnorm(n_sim, mean = mu_p, sd = sigma_p)

  tibble(
    VaR = -quantile(sim_returns, probs = 1 - conf),
    ES = -mean(sim_returns[sim_returns <= quantile(sim_returns, probs = 1 - conf)])
  )
}

mc_results <- bind_rows(
  simulate_portfolio_var(port_returns$`Equal-Weight`) |> mutate(Portfolio = "Equal-Weight"),
  simulate_portfolio_var(port_returns$GMV) |> mutate(Portfolio = "GMV"),
  simulate_portfolio_var(port_returns$`Max Sharpe`) |> mutate(Portfolio = "Max Sharpe")
) |> relocate(Portfolio)

data.frame(mc_results |> mutate(across(c(VaR, ES), ~ round(.x, 4))))
```

**Historical vs. Monte Carlo Distribution**
```{r}
# !VISUALIZATION
# historical vs monte carlo dist
```

#### Stress Testing

```{r}
stress_scenarios <- list(
  "Market-Wide Crash (-20%)"   = rep(-0.20, n_assets),
  "BTC Crash (-50%), rest flat" = c(BDO = 0, ICT = 0, SM = 0, AAPL = 0, MSFT = 0, BTC = -0.50),
  "PH Market Selloff (-15% local, flat abroad)" =
    c(BDO = -0.15, ICT = -0.15, SM = -0.15, AAPL = 0, MSFT = 0, BTC = 0),
  "Global Equity Correction (-10% equities, BTC -25%)" =
    c(BDO = -0.10, ICT = -0.10, SM = -0.10, AAPL = -0.10, MSFT = -0.10, BTC = -0.25)
)

stress_test <- function(weights, scenario) {
  sum(weights * scenario)
}

stress_results <- map_dfr(names(stress_scenarios), function(scenario_name) {
  shock <- stress_scenarios[[scenario_name]]
  tibble(
    Scenario = scenario_name,
    `Equal-Weight` = stress_test(eq_weights, shock),
    GMV = stress_test(gmv_weights, shock),
    `Max Sharpe` = stress_test(max_sharpe_weights, shock)
  )
})

data.frame(stress_results |> mutate(across(where(is.numeric), ~ scales::percent(.x, accuracy = 0.1))))
```

**Stress Testing Chart**
```{r}
# !VISUALIZATION
# stress testing chart
```

## Descriptive Statistics

# Executive Recommendation


# References
