---
title: "main"
output: html_document
date: "2026-08-14"
---
# Case 2 - Portfolio Management and Risk Consultation
## Group Alpha

- Aragon, Enrique Miguel Moreno
- Cardiño, Joaquin Carlo Esteban
- Carlos, Jay Michael Tenorio
- Chua, Erin Gabrielle Tan

This is the code associated with a consulting report on a portfolio 
containing a mixture of assets, specifically the Philippine stocks BDO, ICT, and SM,
the international stocks AAPL and MSFT, and the cryptocurrency BTC. We use SPY
as the primary benchmark to compare against.

# Setup

Initialize packages.

```{r}
pacman::p_load(dplyr, ggplot2, quantmod, tidyverse, tidyquant, lubridate, readxl, timetk, zoo, httr, jsonlite)
```

# Data Acquisition

First, we pull the ticker data from Yahoo! Finance and PSE Edge.
```{r}
DATE_FROM <- "2020-01-02"
DATE_FROM_MDY <- "01-02-2020"

DATE_TO <- "2025-12-31"
DATE_TO_MDY <- "12-31-2025"
```

Many PSE stocks are no longer offered on Yahoo! Finance, so we use another free
alternative called PSE Edge.

```{r}
pse_pull <- function(cmpy_id, security_id, start = DATE_FROM_MDY, end = DATE_TO_MDY) {
  resp <- POST("https://edge.pse.com.ph/common/DisclosureCht.ax",
               body = list(cmpy_id = cmpy_id, security_id = security_id,
                           startDate = start, endDate = end),   # MM-DD-YYYY
               encode = "json",
               add_headers(
                 `X-Requested-With` = "XMLHttpRequest",
                 Referer = "https://edge.pse.com.ph/companyPage/stockData.do",
                 Origin = "https://edge.pse.com.ph",
                 Accept = "application/json, text/javascript, */*; q=0.01"))
  df <- fromJSON(content(resp, "text", encoding = "UTF-8"))$chartData
  df$date <- as.Date(df$CHART_DATE, format = "%b %d, %Y 00:00:00")
  df[order(df$date), c("date", "OPEN", "HIGH", "LOW", "CLOSE")]
}

yahoo_pull <- function(sym, start = DATE_FROM, end = DATE_TO) {
  tq_get(sym, from = start, to = end) |>
    transmute(date = as.Date(date), !!sym := round(adjusted, 4))
}

Sys.sleep(1); bdo <- pse_pull(260, 468); Sys.sleep(1)
Sys.sleep(1); ict <- pse_pull(83, 142);  Sys.sleep(1)
Sys.sleep(1); sm  <- pse_pull(599, 520); Sys.sleep(1)

prices <- list(
  BDO = bdo |> rename(BDO = CLOSE),
  ICT = ict |> rename(ICT = CLOSE),
  SM  = sm  |> rename(SM = CLOSE),
  AAPL = yahoo_pull("AAPL"), MSFT = yahoo_pull("MSFT"), BTC = yahoo_pull("BTC-USD"), SPY = yahoo_pull("SPY")
) |> reduce(inner_join, by = "date") |> arrange(date)

dir.create("data", showWarnings = FALSE)
write.csv(prices, "data/prices.csv", row.names = FALSE)
save(prices, file = "data/prices.RData")
```

We also get the 91-day Treasury bill as the risk-free rate.
We have to manually download it from https://data.bettergov.ph/datasets/12/resources/159
because the website blocks automated downloads.

```{r}
tbill91 <- unlist(lapply(2020:2025, function(yr) {
  gs <- read_excel("data/gs_rates_monthly.xlsx", sheet = as.character(yr), skip = 4)
  row_91 <- gs[which(grepl("91-day", as.character(gs[[2]]))), ]   # label in col 2
  as.numeric(unlist(row_91[1, seq(3, 25, by = 2)]))               # C,E,G,...,Y
}))
```

We then clean the Treasury bill rates by dropping the blank months and
computing the annual and daily risk-free rate.

```{r}
tbill91_pct <- na.omit(tbill91)                # drops blank / no-auction months
rf_annual <- mean(tbill91_pct) / 100           # ~4% expected
rf_daily  <- (1 + rf_annual)^(1 / 252) - 1     # use in Sharpe / frontier
```

# Exploratory Financial Analysis


# Portfolio Construction


# Portfolio Optimization


# Portfolio Risk Assessment


# Executive Recommendation


# References
